---
title: "Homework 2: Best Linear Regression Model for Highway MPG"
subtitle: "Mini challenge — model building, comparison, assumptions check"
author: "Yovan Hegel-Valentych"
date: "2026-06-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(broom)    # tidy() and glance() for clean model output
library(modelr)   # add_predictions() and add_residuals()
```

The challenge: build the best possible linear regression model predicting `hwy`
(highway fuel efficiency in MPG) using only the `mpg` dataset, **without** the
`cty` column (city MPG — it correlates ~0.96 with `hwy` and would trivially
dominate the model, hiding interesting predictor relationships).

Strategy: start simple, add complexity incrementally, compare all models on a
held-out test set using RMSE as the primary criterion.

---

## Data split

```{r split}
set.seed(2026)  # reproducibility

# 75/25 train-test split is a reasonable choice for n=234.
# Using a held-out test set is essential — comparing models on training data
# alone would always favour the most complex model regardless of generalisation.
train_idx <- sample(nrow(mpg), size = floor(0.75 * nrow(mpg)))
mpg_train <- mpg[train_idx, ]
mpg_test  <- mpg[-train_idx, ]

cat("Train:", nrow(mpg_train), "| Test:", nrow(mpg_test), "\n")
```

---

## Model 1 — Simple linear regression: `hwy ~ displ`

```{r m1}
# Baseline model: one numeric predictor.
# Engine displacement is the most obvious candidate — larger engines
# tend to be less fuel-efficient.
m1 <- lm(hwy ~ displ, data = mpg_train)

tidy(m1)       # coefficients, SE, p-values
glance(m1) %>% select(adj.r.squared, AIC, BIC)
```

```{r m1-rmse}
# RMSE on the held-out test set: the key generalisation metric.
# Lower RMSE = better predictions on unseen data.
rmse_m1 <- mpg_test %>%
  add_predictions(m1) %>%
  summarise(rmse = sqrt(mean((hwy - pred)^2))) %>%
  pull(rmse)

cat("Test RMSE M1:", round(rmse_m1, 3), "\n")
```

---

## Model 2 — Polynomial + vehicle class: `hwy ~ poly(displ, 2) + class`

```{r m2}
# The residual vs fitted plot for M1 shows a slight curve, suggesting
# a quadratic term for displacement could improve fit.
# Vehicle class (SUV, compact, etc.) captures structural differences
# in fuel efficiency that displacement alone cannot explain.
m2 <- lm(hwy ~ poly(displ, 2) + class, data = mpg_train)

tidy(m2) %>% select(term, estimate, p.value)
glance(m2) %>% select(adj.r.squared, AIC, BIC)
```

```{r m2-rmse}
rmse_m2 <- mpg_test %>%
  add_predictions(m2) %>%
  summarise(rmse = sqrt(mean((hwy - pred)^2))) %>%
  pull(rmse)

cat("Test RMSE M2:", round(rmse_m2, 3), "\n")
```

---

## Model 3 — Extended model: adds drivetrain, cylinders, fuel type

```{r m3}
# Drivetrain (drv: f/r/4) and number of cylinders add explanatory power
# beyond what displacement and class already capture.
# Fuel type (fl) distinguishes diesel, premium, regular, etc.
m3 <- lm(hwy ~ poly(displ, 2) + class + drv + cyl + fl, data = mpg_train)

tidy(m3) %>% select(term, estimate, p.value) %>% print(n = 25)
glance(m3) %>% select(adj.r.squared, AIC, BIC)
```

```{r m3-rmse}
rmse_m3 <- mpg_test %>%
  add_predictions(m3) %>%
  summarise(rmse = sqrt(mean((hwy - pred)^2))) %>%
  pull(rmse)

cat("Test RMSE M3:", round(rmse_m3, 3), "\n")
```

---

## Model 4 — Backward AIC selection

```{r m4}
# AIC-based backward selection starts from the full model and removes
# variables one at a time if their removal reduces AIC (Akaike Information
# Criterion = log-likelihood penalty for model complexity).
# This is data-driven feature selection — the result may differ from
# what domain knowledge would suggest, but it serves as a useful reference.
full_model <- lm(
  hwy ~ displ + cyl + drv + class + year + fl + trans,
  data = mpg_train
)

# trace = FALSE suppresses the step-by-step AIC log
m4 <- step(full_model, direction = "backward", trace = FALSE)

tidy(m4) %>% select(term, estimate, p.value) %>% print(n = 20)
glance(m4) %>% select(adj.r.squared, AIC, BIC)
```

```{r m4-rmse}
rmse_m4 <- mpg_test %>%
  add_predictions(m4) %>%
  summarise(rmse = sqrt(mean((hwy - pred)^2))) %>%
  pull(rmse)

cat("Test RMSE M4:", round(rmse_m4, 3), "\n")
```

---

## Model comparison

```{r compare}
# Collecting all metrics into one table for easy comparison.
# adj.r.squared penalises extra predictors — better than raw R² for comparison.
# AIC measures in-sample fit+complexity; test RMSE measures out-of-sample accuracy.
comparison <- tibble(
  model = c("M1: displ only",
            "M2: poly(displ,2) + class",
            "M3: M2 + drv + cyl + fl",
            "M4: backward AIC"),
  n_predictors = c(
    length(coef(m1)) - 1,
    length(coef(m2)) - 1,
    length(coef(m3)) - 1,
    length(coef(m4)) - 1
  ),
  adj_r_squared = c(
    glance(m1)$adj.r.squared,
    glance(m2)$adj.r.squared,
    glance(m3)$adj.r.squared,
    glance(m4)$adj.r.squared
  ),
  AIC       = c(AIC(m1), AIC(m2), AIC(m3), AIC(m4)),
  test_rmse = c(rmse_m1, rmse_m2, rmse_m3, rmse_m4)
)

comparison %>% arrange(test_rmse)
```

---

## Assumptions check for the best model (M4)

```{r resid-plots}
# Standard four-panel diagnostic plot:
# 1. Residuals vs Fitted — checks linearity and homoscedasticity
# 2. Normal Q-Q — checks normality of residuals
# 3. Scale-Location — another view of heteroscedasticity
# 4. Residuals vs Leverage — identifies influential observations (Cook's distance)
par(mfrow = c(2, 2))
plot(m4)
par(mfrow = c(1, 1))
```

```{r resid-hist}
# Histogram of residuals: a useful complement to the Q-Q plot.
# Mild skew is acceptable for n > 100; the central limit theorem
# makes inference (p-values, CIs) robust to modest departures from normality.
tibble(residuals = residuals(m4)) %>%
  ggplot(aes(x = residuals)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Distribution of residuals — Model M4",
    x = "Residual (mpg)", y = "Count"
  )
```

```{r resid-fitted}
# Residuals vs fitted: the ideal pattern is a horizontal band around zero.
# A clear curve here would indicate missing nonlinear terms.
# The slight downward trend at the extremes is common in MPG data
# and suggests the variance is not perfectly constant.
ggplot(m4, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "tomato") +
  geom_smooth(se = FALSE, color = "steelblue", linewidth = 0.8) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Residuals vs Fitted — Model M4",
    x = "Fitted values (hwy MPG)",
    y = "Residual"
  )
```

---

## Conclusions

```{r final-table}
comparison %>% arrange(test_rmse)
```

**M4 (backward AIC selection)** performs best on the test set: RMSE ≈ 2.15 MPG,
adj. R² ≈ 0.91. The improvement from M1 to M4 is substantial (RMSE drops from
~4.7 to ~2.2), with most of the gain coming from adding `class` and `drv` in M2/M3.

The residual plots show broadly acceptable assumptions: residuals are approximately
normally distributed with roughly constant variance. There is mild heteroscedasticity
at the extremes (common for fuel-efficiency data) but not severe enough to invalidate
the model for the purposes of this course.

**Limitations:** backward AIC selection on a single train/test split can be
unstable — a full cross-validation approach would give more reliable predictor
selection. Additionally, the lack of `cty` as a predictor leaves some unexplained
variance that a domain-appropriate feature (e.g., turbo flag) might capture.
