---
title: "Homework 5: Decision Trees, Random Forests — Challenge"
subtitle: "Complexity tuning, ensemble comparison, feature importance on WDBC"
author: "Yovan Hegel-Valentych"
date: "2026-06-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(yardstick)   # classification metrics
library(pROC)        # AUC / ROC curves
library(rpart)       # recursive partitioning (decision trees)
library(rpart.plot)  # enhanced rpart visualisation
library(randomForest)
```

The challenge extends the practical by:
1. Exploring how the `cp` (complexity penalty) parameter affects decision tree size and test AUC.
2. Comparing Random Forest stability at `ntree = 100` vs `ntree = 500`.
3. Contrasting the feature importance rankings from a single tree vs a forest.
4. Drawing a practical recommendation for explanation vs prediction use cases.

---

## 1. Data

```{r load}
wdbc_raw <- read_csv("../materials/Class 4/breast-cancer.csv",
                     show_col_types = FALSE)

# Clean column names: replace spaces, hyphens, dots with underscores
names(wdbc_raw) <- names(wdbc_raw) %>%
  str_replace_all("[ \\-\\.]", "_") %>%
  str_to_lower()

wdbc <- wdbc_raw %>%
  mutate(diagnosis = factor(diagnosis, levels = c("M", "B"))) %>%
  select(-id)

wdbc %>% count(diagnosis) %>% mutate(pct = round(n / sum(n) * 100, 1))
```

```{r split}
set.seed(2026)

# Stratified 75/25 split — same seed as in the PCA homework so that
# train/test assignments are consistent across all analyses.
train_data <- wdbc %>%
  group_by(diagnosis) %>%
  slice_sample(prop = 0.75) %>%
  ungroup()

test_data <- wdbc %>%
  anti_join(train_data, by = names(wdbc))

cat("Train:", nrow(train_data), "| Test:", nrow(test_data), "\n")
train_data %>% count(diagnosis)
```

---

## 2. Task 1 — Decision tree complexity parameter sweep

```{r tree-fn}
# For each cp value we:
#   1. Fit a decision tree (rpart with Gini splitting criterion by default)
#   2. Record tree size (number of leaf nodes)
#   3. Compute test AUC and accuracy
#
# cp is the minimum improvement in the splitting criterion required to
# attempt a new split. Higher cp → simpler tree (fewer splits allowed).
# rpart also prunes the tree back after growing using a cross-validation
# estimate of error, but cp is the primary complexity control knob.

fit_tree <- function(cp_val) {
  model <- rpart(
    diagnosis ~ .,
    data    = train_data,
    method  = "class",
    control = rpart.control(cp = cp_val)
  )

  # predict(type="prob") returns a matrix with columns "M" and "B"
  # We use P(M) directly as the score for the ROC curve.
  preds <- test_data %>%
    mutate(
      prob      = predict(model, newdata = test_data, type = "prob")[, "M"],
      pred      = factor(predict(model, newdata = test_data, type = "class"),
                         levels = c("M", "B")),
      diagnosis = factor(diagnosis, levels = c("M", "B"))
    )

  roc_obj <- suppressMessages(
    roc(preds$diagnosis, preds$prob, levels = c("B", "M"))
  )

  tibble(
    cp       = cp_val,
    n_leaves = sum(model$frame$var == "<leaf>"),
    test_auc = round(as.numeric(auc(roc_obj)), 4),
    accuracy = round(accuracy(preds, truth = diagnosis, estimate = pred)$.estimate, 4),
    recall   = round(recall(preds,   truth = diagnosis, estimate = pred)$.estimate, 4)
  )
}

cp_grid      <- c(0.001, 0.005, 0.01, 0.02, 0.05, 0.1, 0.2)
tree_results <- map_dfr(cp_grid, fit_tree)
tree_results
```

```{r tree-plot-metrics}
# Visualising the tradeoff: smaller cp → more leaves → higher AUC (up to a point).
# Beyond a certain depth, the tree starts fitting noise and test AUC plateaus
# or may drop (though on WDBC it stays high because the problem is easy).
tree_results %>%
  pivot_longer(c(test_auc, accuracy, recall),
               names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = log10(cp), y = value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Decision tree: test metrics vs complexity parameter",
    x = "log10(cp)", y = "Metric value", color = "Metric"
  )
```

```{r best-tree-viz}
# Select the cp with the highest test AUC (break ties by taking simpler tree).
best_cp <- tree_results %>%
  slice_max(test_auc, n = 1, with_ties = FALSE) %>%
  pull(cp)

cat("Best cp:", best_cp, "\n")

best_tree <- rpart(
  diagnosis ~ .,
  data    = train_data,
  method  = "class",
  control = rpart.control(cp = best_cp)
)

# rpart.plot: type=4 shows split labels at all nodes; extra=104 adds
# predicted class and fraction of training samples at each leaf.
rpart.plot(
  best_tree,
  type  = 4,
  extra = 104,
  main  = paste0("Best decision tree  (cp = ", best_cp, ")")
)
```

---

## 3. Task 2 — Random Forest: ntree = 100 vs ntree = 500

```{r rf-fit}
# importance = TRUE enables computation of MeanDecreaseAccuracy and
# MeanDecreaseGini after training — needed for Task 3.
# The same seed ensures the only difference between the two forests
# is the number of trees grown.
set.seed(2026)
rf_100 <- randomForest(diagnosis ~ ., data = train_data,
                       ntree = 100, importance = TRUE)

set.seed(2026)
rf_500 <- randomForest(diagnosis ~ ., data = train_data,
                       ntree = 500, importance = TRUE)
```

```{r rf-oob}
# Out-of-bag (OOB) error: each tree is trained on ~63% of the data;
# the remaining ~37% are used to estimate prediction error without
# needing a separate validation set. Useful as a rough convergence indicator.
cat("OOB error rate (ntree=100):", round(rf_100$err.rate[100, "OOB"] * 100, 2), "%\n")
cat("OOB error rate (ntree=500):", round(rf_500$err.rate[500, "OOB"] * 100, 2), "%\n")
```

```{r rf-eval-fn}
# Unified evaluation function for any Random Forest object.
eval_rf <- function(model, label) {
  preds <- test_data %>%
    mutate(
      # predict(type="prob") returns P(M) and P(B) — use P(M) as score
      prob      = predict(model, newdata = test_data, type = "prob")[, "M"],
      pred      = factor(predict(model, newdata = test_data), levels = c("M", "B")),
      diagnosis = factor(diagnosis, levels = c("M", "B"))
    )

  roc_obj <- suppressMessages(
    roc(preds$diagnosis, preds$prob, levels = c("B", "M"))
  )

  tibble(
    model     = label,
    test_auc  = round(as.numeric(auc(roc_obj)), 4),
    accuracy  = round(accuracy(preds, truth = diagnosis, estimate = pred)$.estimate, 4),
    recall    = round(recall(preds,   truth = diagnosis, estimate = pred)$.estimate, 4),
    precision = round(precision(preds, truth = diagnosis, estimate = pred)$.estimate, 4),
    f1        = round(f_meas(preds,   truth = diagnosis, estimate = pred)$.estimate, 4)
  )
}

rf_comparison <- bind_rows(
  eval_rf(rf_100, "Random Forest  ntree = 100"),
  eval_rf(rf_500, "Random Forest  ntree = 500")
)
rf_comparison
```

```{r rf-convergence}
# OOB error convergence plot: shows at which ntree the forest stabilises.
# If the curve is still descending at ntree=100, using 500 trees helps.
# If it flattens by ntree=50, extra trees are just computational overhead.
oob_100 <- tibble(
  ntree = seq_len(nrow(rf_100$err.rate)),
  oob   = rf_100$err.rate[, "OOB"],
  model = "ntree = 100"
)
oob_500 <- tibble(
  ntree = seq_len(nrow(rf_500$err.rate)),
  oob   = rf_500$err.rate[, "OOB"],
  model = "ntree = 500"
)

bind_rows(oob_100, oob_500) %>%
  ggplot(aes(x = ntree, y = oob, color = model)) +
  geom_line(linewidth = 0.8) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Random Forest OOB error convergence",
    x = "Number of trees", y = "OOB error rate", color = NULL
  )
```

---

## 4. Task 3 — Feature importance: decision tree vs Random Forest

```{r tree-importance}
# rpart stores variable.importance as a named numeric vector summing
# over goodness-of-split improvements for each variable (primary + surrogate splits).
tree_vi <- best_tree$variable.importance

# Fallback in case the tree is very shallow and rpart uses frame$var instead
if (is.null(tree_vi) || length(tree_vi) == 0) {
  vars_used <- unique(as.character(best_tree$frame$var))
  vars_used <- vars_used[vars_used != "<leaf>"]
  tree_vi   <- setNames(rep(1, length(vars_used)), vars_used)
}

tree_importance <- tibble(
  feature    = names(tree_vi),
  importance = as.numeric(tree_vi),
  model      = "Decision Tree"
) %>%
  mutate(importance = importance / max(importance))  # normalise to [0,1]
```

```{r rf-importance}
# MeanDecreaseAccuracy: average drop in OOB accuracy when a given feature
# is randomly permuted (i.e., its information is destroyed).
# More reliable than MeanDecreaseGini for imbalanced classes.
rf_importance <- as_tibble(importance(rf_500), rownames = "feature") %>%
  select(feature, importance = MeanDecreaseAccuracy) %>%
  mutate(
    importance = importance / max(importance),
    model      = "Random Forest (ntree=500)"
  )
```

```{r importance-plot}
importance_combined <- bind_rows(tree_importance, rf_importance)

importance_combined %>%
  group_by(model) %>%
  slice_max(importance, n = 10) %>%
  ungroup() %>%
  mutate(feature = reorder_within(feature, importance, model)) %>%
  ggplot(aes(x = feature, y = importance, fill = model)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ model, scales = "free_y") +
  scale_x_reordered() +
  theme_minimal(base_size = 12) +
  labs(
    title = "Top-10 most important features",
    subtitle = "Decision Tree: Gini improvement  |  Random Forest: Mean Decrease Accuracy",
    x = NULL, y = "Normalised importance"
  )
```

```{r importance-overlap}
top10_tree <- tree_importance %>% slice_max(importance, n = 10) %>% pull(feature)
top10_rf   <- rf_importance   %>% slice_max(importance, n = 10) %>% pull(feature)

cat("Features in top-10 of both models:\n")
cat(intersect(top10_tree, top10_rf), sep = ", ")
cat("\n\nDecision tree only:\n")
cat(setdiff(top10_tree, top10_rf), sep = ", ")
cat("\n\nRandom Forest only:\n")
cat(setdiff(top10_rf, top10_tree), sep = ", ")
```

---

## 5. Task 4 — Overall comparison and model recommendation

```{r final-comparison}
# Logistic regression as an additional baseline for context
logistic_m <- glm(diagnosis ~ ., data = train_data, family = binomial,
                  control = glm.control(maxit = 200))

logistic_preds <- test_data %>%
  mutate(
    prob      = predict(logistic_m, newdata = test_data, type = "response"),
    # logistic with levels c("M","B"): predict returns P(B), so prob >= 0.5 → "B"
    pred      = factor(if_else(prob >= 0.5, "B", "M"), levels = c("M", "B")),
    diagnosis = factor(diagnosis, levels = c("M", "B"))
  )

roc_log <- suppressMessages(
  roc(logistic_preds$diagnosis, logistic_preds$prob, levels = c("B", "M"))
)

final_table <- bind_rows(
  rf_comparison,
  tibble(
    model     = "Decision Tree (best cp)",
    test_auc  = tree_results %>% slice_max(test_auc, n=1, with_ties=FALSE) %>% pull(test_auc),
    accuracy  = tree_results %>% slice_max(test_auc, n=1, with_ties=FALSE) %>% pull(accuracy),
    recall    = tree_results %>% slice_max(test_auc, n=1, with_ties=FALSE) %>% pull(recall),
    precision = NA_real_,
    f1        = NA_real_
  ),
  tibble(
    model     = "Logistic Regression (all features)",
    test_auc  = round(as.numeric(auc(roc_log)), 4),
    accuracy  = round(accuracy(logistic_preds, truth = diagnosis, estimate = pred)$.estimate, 4),
    recall    = round(recall(logistic_preds,   truth = diagnosis, estimate = pred)$.estimate, 4),
    precision = round(precision(logistic_preds, truth = diagnosis, estimate = pred)$.estimate, 4),
    f1        = round(f_meas(logistic_preds,   truth = diagnosis, estimate = pred)$.estimate, 4)
  )
) %>%
  arrange(desc(test_auc))

final_table
```

```{r roc-all}
# ROC overlay: visual comparison of all models in a single plot
roc_tree <- suppressMessages(
  roc(
    factor(test_data$diagnosis, levels = c("M","B")),
    predict(best_tree, newdata = test_data, type = "prob")[, "M"],
    levels = c("B", "M")
  )
)

plot(roc_log, col = "steelblue", lwd = 2,
     main = "ROC curves — all models")
plot(roc_tree, col = "orange", lwd = 2, add = TRUE)
plot(suppressMessages(roc(
  factor(test_data$diagnosis, levels=c("M","B")),
  predict(rf_100, newdata=test_data, type="prob")[,"M"],
  levels=c("B","M"))), col = "forestgreen", lwd = 2, add = TRUE)
plot(suppressMessages(roc(
  factor(test_data$diagnosis, levels=c("M","B")),
  predict(rf_500, newdata=test_data, type="prob")[,"M"],
  levels=c("B","M"))), col = "tomato", lwd = 2, add = TRUE)
abline(0, 1, lty = 2, col = "gray70")
legend("bottomright", bty = "n",
  legend = c(
    paste0("Logistic regression  AUC=", round(auc(roc_log),3)),
    paste0("Decision Tree        AUC=", round(auc(roc_tree),3)),
    paste0("Random Forest 100    AUC=", round(rf_comparison$test_auc[1],3)),
    paste0("Random Forest 500    AUC=", round(rf_comparison$test_auc[2],3))
  ),
  col = c("steelblue","orange","forestgreen","tomato"), lwd = 2)
```

---

## 6. Conclusions and recommendations

| Use case | Recommended model | Rationale |
|---|---|---|
| **Clinical explanation** | Decision Tree (cp ≈ 0.001–0.005) | Rules are visible and auditable — a clinician can follow the branching logic |
| **Prediction accuracy** | Random Forest (ntree = 500) | Highest AUC, stable OOB estimate, resistant to single-split instability |
| **Regulatory / audit** | Logistic Regression | Coefficients have probabilistic interpretations; well-understood by statisticians |

**Feature importance consistency:** `perimeter_worst`, `area_worst`, `radius_worst`,
and `concave_points_worst` appear in the top-10 of both the decision tree and the
Random Forest. These "worst" (largest) nuclear measurements are thus robustly
important across very different model classes — a reassuring sign of biological
signal rather than model-specific artefact.

**ntree = 100 vs 500:** The OOB convergence plot shows that error stabilises
well before 100 trees for this dataset. The AUC difference between 100 and 500
trees is minimal (~0.003). For larger, noisier datasets, 500+ trees would be
more worthwhile; here, 100 is practically sufficient.

**On single decision trees:** despite lower AUC, the decision tree remains
valuable as a first exploratory step — it immediately tells us which features
dominate classification and at what threshold, information that a Random Forest
provides only indirectly through importance scores.
