---
title: "Homework 4: PCA and Dimensionality Reduction — Extension"
subtitle: "How many principal components are enough for classification on WDBC?"
author: "Yovan Hegel-Valentych"
date: "2026-06-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(yardstick)
library(pROC)
```

The practical covered the mechanics of PCA and showed that dimensionality can be
reduced while preserving the bulk of variance. This extension asks a quantitative
question: **how many principal components are actually needed to maintain
classification performance close to the full-feature baseline?**

The answer has practical importance in biomedical data (high-dimensional omics,
imaging features) where fitting models on 30+ raw features is often unnecessary,
unstable, or uninterpretable.

---

## 1. Load and prepare data

```{r load}
wdbc_raw <- read_csv("../materials/Class 4/breast-cancer.csv",
                     show_col_types = FALSE)

# Sanitise column names: spaces, hyphens, and dots → underscores, all lowercase.
# This avoids backtick-quoting headaches in formulas and dplyr pipes.
names(wdbc_raw) <- names(wdbc_raw) %>%
  str_replace_all("[ \\-\\.]", "_") %>%
  str_to_lower()

wdbc <- wdbc_raw %>%
  # Positive class first — yardstick and pROC conventions
  mutate(diagnosis = factor(diagnosis, levels = c("M", "B"))) %>%
  select(-id)  # ID is a row identifier, not a predictor

glimpse(wdbc)
wdbc %>% count(diagnosis) %>% mutate(pct = round(n / sum(n) * 100, 1))
```

---

## 2. Stratified train / test split

```{r split}
set.seed(2026)

# Stratified sampling within each class ensures both sets have ~37% M / 63% B.
# Using anti_join on all columns to extract the exact complement of train rows.
wdbc_train <- wdbc %>%
  group_by(diagnosis) %>%
  slice_sample(prop = 0.75) %>%
  ungroup()

wdbc_test <- wdbc %>%
  anti_join(wdbc_train, by = names(wdbc))

cat("Train:", nrow(wdbc_train), "| Test:", nrow(wdbc_test), "\n")
wdbc_train %>% count(diagnosis)
```

```{r matrices}
# Separate feature matrices and target vectors for PCA and downstream modelling.
x_train <- wdbc_train %>% select(-diagnosis) %>% as.matrix()
y_train  <- wdbc_train$diagnosis
x_test   <- wdbc_test  %>% select(-diagnosis) %>% as.matrix()
y_test   <- wdbc_test$diagnosis
```

---

## 3. Standardise features

```{r scale}
# PCA is sensitive to scale: a feature with large variance dominates the
# principal components regardless of its biological relevance.
# Standardisation (mean = 0, sd = 1) removes this bias.
#
# IMPORTANT: mean and SD are computed ONLY on the training set and then
# applied to the test set. Using test statistics here would constitute
# data leakage.
train_means <- colMeans(x_train)
train_sds   <- apply(x_train, 2, sd)

x_train_sc <- scale(x_train, center = train_means, scale = train_sds)
x_test_sc  <- scale(x_test,  center = train_means, scale = train_sds)
```

---

## 4. Fit PCA on training data

```{r pca}
# prcomp() returns: rotation matrix (loadings), x (scores), sdev (std deviations).
# center = FALSE and scale. = FALSE because we already standardised manually.
pca_fit <- prcomp(x_train_sc, center = FALSE, scale. = FALSE)

var_exp <- pca_fit$sdev^2 / sum(pca_fit$sdev^2)
cum_var <- cumsum(var_exp)

n_pc_90 <- which(cum_var >= 0.90)[1]
n_pc_95 <- which(cum_var >= 0.95)[1]
cat("PCs to explain 90% variance:", n_pc_90, "\n")
cat("PCs to explain 95% variance:", n_pc_95, "\n")
```

```{r scree}
# Scree plot with cumulative variance overlay.
# The "elbow" in the scree (bar chart) often suggests a natural cutoff,
# but the cumulative line with a 90% threshold is more objective.
tibble(pc = seq_along(var_exp), ve = var_exp, cv = cum_var) %>%
  filter(pc <= 20) %>%
  ggplot(aes(x = pc)) +
  geom_col(aes(y = ve), fill = "steelblue", alpha = 0.7) +
  geom_line(aes(y = cv), color = "tomato", linewidth = 1.2) +
  geom_point(aes(y = cv), color = "tomato", size = 2.5) +
  geom_hline(yintercept = 0.90, linetype = "dashed", color = "gray50") +
  annotate("text", x = n_pc_90 + 0.5, y = 0.87,
           label = paste0("90% @ PC", n_pc_90), hjust = 0, size = 3.5) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Scree plot with cumulative explained variance",
    subtitle = "Bars = individual PC contribution; line = cumulative sum",
    x = "Principal Component", y = "Proportion of variance explained"
  )
```

---

## 5. PC1 vs PC2 scatter

```{r biplot}
# PC1 and PC2 together explain the largest possible fraction of variance
# in two dimensions. A clean separation here suggests the two classes
# differ primarily along the directions of maximum variance — a good sign
# for any downstream linear classifier.
as_tibble(pca_fit$x[, 1:2]) %>%
  mutate(diagnosis = y_train) %>%
  ggplot(aes(x = PC1, y = PC2, color = diagnosis)) +
  geom_point(alpha = 0.7, size = 2.2) +
  theme_minimal(base_size = 13) +
  labs(
    title = "PCA score plot: PC1 vs PC2 (training set)",
    x = paste0("PC1 (", round(var_exp[1] * 100, 1), "% variance)"),
    y = paste0("PC2 (", round(var_exp[2] * 100, 1), "% variance)"),
    color = "Diagnosis"
  )
```

---

## 6. Classification performance vs number of PCs

```{r pca-logistic}
# For each candidate number of PCs:
#   1. Project train and test data onto the first k PCs
#   2. Fit logistic regression on the projected training data
#   3. Predict on the projected test data
#   4. Compute AUC, accuracy, and recall
#
# Note on GLM warnings: "fitted probabilities numerically 0 or 1 occurred"
# indicates perfect or near-perfect separation of classes in the PC space.
# This is expected for WDBC at k >= 5 — the classes ARE well-separated.
# The GLM still converges and produces meaningful AUC, though coefficient
# SEs are inflated (not an issue here since we care about prediction).

fit_pca_logistic <- function(n_components) {
  # Project onto first k PCs using the PCA fit from training data only
  train_pca <- as_tibble(pca_fit$x[, 1:n_components]) %>%
    mutate(diagnosis = y_train)

  test_pca <- as_tibble(
    predict(pca_fit, newdata = x_test_sc)[, 1:n_components]
  ) %>%
    mutate(diagnosis = y_test)

  # Fit logistic regression in the reduced PC space
  model <- suppressWarnings(
    glm(diagnosis ~ ., data = train_pca, family = binomial,
        control = glm.control(maxit = 200))
  )

  # predict(type="response") returns P(B) because levels = c("M","B")
  # Correct threshold: prob >= 0.5 → "B"
  preds <- test_pca %>%
    mutate(
      prob      = predict(model, newdata = test_pca, type = "response"),
      pred      = factor(if_else(prob >= 0.5, "B", "M"), levels = c("M", "B")),
      diagnosis = factor(diagnosis, levels = c("M", "B"))
    )

  roc_obj <- suppressMessages(
    roc(preds$diagnosis, preds$prob, levels = c("B", "M"))
  )

  tibble(
    n_components = n_components,
    auc          = round(as.numeric(auc(roc_obj)), 4),
    accuracy     = round(accuracy(preds, truth = diagnosis, estimate = pred)$.estimate, 4),
    recall       = round(recall(preds,   truth = diagnosis, estimate = pred)$.estimate, 4),
    precision    = round(precision(preds, truth = diagnosis, estimate = pred)$.estimate, 4),
    f1           = round(f_meas(preds,   truth = diagnosis, estimate = pred)$.estimate, 4)
  )
}

# Test several candidate sizes including the 90%-variance cutoff
component_grid <- c(2, 5, 7, 10, 15, n_pc_90, 30)
pc_results <- map_dfr(unique(sort(component_grid)), fit_pca_logistic)

pc_results
```

---

## 7. Visual comparison

```{r compare-plot}
# Performance as a function of the number of components.
# We expect AUC to plateau quickly — the point at which adding more PCs
# no longer improves the test AUC is the practical dimensionality cutoff.
pc_results %>%
  pivot_longer(c(auc, accuracy, recall, f1),
               names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = n_components, y = value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Classification metrics vs number of PCA components",
    x = "Number of components", y = "Metric value",
    color = "Metric"
  )
```

---

## 8. Answers to discussion questions

```{r best-n}
# Which number of components gives the best AUC?
pc_results %>% slice_max(auc, n = 3)
```

**Does PCA reduce dimensionality without major performance loss?**
Yes. Even 2–5 principal components (out of 30 original features) achieve AUC > 0.97,
close to the full-feature model. This is a direct consequence of the high
collinearity among the WDBC features — most variance is concentrated in the
first few PCs.

**How many components are enough?**
`r n_pc_90` components explain 90% of variance and deliver near-maximal AUC.
Beyond that, incremental gains are negligible while the model becomes harder
to interpret.

**PCA for prediction vs visualisation?**
- For **visualisation**: 2 PCs are sufficient. The score plot (Section 5) shows
  near-perfect class separation already in 2D.
- For **prediction**: 5–10 PCs give stable, high-AUC models. Beyond that,
  using more components risks capturing noise rather than signal — the logistic
  GLM may fail to converge due to perfect separation in high-dimensional PC space.

```{r summary-table}
pc_results %>% arrange(desc(auc))
```
