---
title: "Homework 1: Cleaning and Tidying Messy Penguin Data"
subtitle: "Tidyverse workflow — inspect, clean, filter, mutate, pivot, join, summarise"
author: "Yovan Hegel-Valentych"
date: "2026-06-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
```

The goal of this homework is to practice the full tidyverse pipeline on a deliberately
messy version of the Palmer Penguins dataset: from raw inspection to a clean,
summarised analytical result.

---

## Read data

```{r load}
# Both files live one level up inside the course materials folder.
# The penguin observations have 11 columns including quality flags and batch IDs
# that are absent from the original palmerpenguins package.
data_penguins   <- read_csv("../materials/Class 1/messy_penguins.csv")
metadata_islands <- read_csv("../materials/Class 1/island_metadata.csv")
```

---

## Step 1: Inspect the data

```{r inspect-dims}
# Basic dimensions — good starting point before any transformation
nrow(data_penguins)
ncol(data_penguins)
names(data_penguins)
```

```{r inspect-summary}
# summary() gives a quick view of ranges and missing counts per column
summary(data_penguins)
```

```{r inspect-missing}
# Counting NAs per column explicitly — easier to scan than summary() output
data_penguins %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") %>%
  filter(n_missing > 0)
```

```{r inspect-categoricals}
# Checking unique values in categorical columns reveals the messy parts:
# species and island have mixed letter case (e.g. "ADELIE" vs "Adelie"),
# sex has mixed case too ("FEMALE" vs "female")
data_penguins %>% count(species)
data_penguins %>% count(island)
data_penguins %>% count(sex)
data_penguins %>% count(observation_quality)
```

**Observations:** the dataset contains mixed-case labels in `species`, `island`,
and `sex`, plus some missing values in body measurements and sex.
The `observation_quality` column distinguishes high, medium, and low-quality records.

---

## Step 2: Clean categorical variables

```{r clean}
# str_trim() removes any accidental leading/trailing whitespace.
# str_to_title() / str_to_lower() standardise letter case so that
# "ADELIE", "Adelie", "adelie" all become the same token.
# drop_na() on the key measurement columns removes rows where we cannot
# compute meaningful statistics — better to lose a few rows than carry NAs
# silently into downstream analyses.
penguins_cleaned_homework <- data_penguins %>%
  mutate(
    species             = str_trim(str_to_title(species)),
    island              = str_trim(str_to_title(island)),
    sex                 = str_trim(str_to_lower(sex)),
    observation_quality = str_trim(str_to_lower(observation_quality))
  ) %>%
  drop_na(species, island, sex,
          bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)

cat("Rows remaining after cleaning:", nrow(penguins_cleaned_homework), "\n")
penguins_cleaned_homework %>% count(sex)
penguins_cleaned_homework %>% count(observation_quality)
```

---

## Step 3: Filter useful observations

```{r filter}
# Retaining only observations that pass a minimum quality threshold,
# have a body mass above 3 kg (removes suspiciously small measurements),
# and come from the two islands with the most data: Biscoe and Dream.
# Torgersen is excluded — too few samples for reliable group statistics.
penguins_filtered <- penguins_cleaned_homework %>%
  filter(
    observation_quality %in% c("high", "medium"),
    body_mass_g > 3000,
    island %in% c("Biscoe", "Dream")
  )

cat("Rows after filtering:", nrow(penguins_filtered), "\n")

# Verify the group sizes are reasonable for downstream aggregation
penguins_filtered %>% count(island, observation_quality)
```

---

## Step 4: Add new variables with `mutate()`

```{r mutate}
# body_mass_kg is just a unit conversion — more readable in plots.
# bill_ratio captures the shape of the bill; morphologically informative
# for species discrimination (longer + shallower vs. shorter + deeper).
# large_penguin is a simple binary flag useful for quick group comparisons.
penguins_mutated <- penguins_filtered %>%
  mutate(
    body_mass_kg  = body_mass_g / 1000,
    bill_ratio    = bill_length_mm / bill_depth_mm,
    large_penguin = if_else(body_mass_g > 4500, "large", "regular")
  )

penguins_mutated %>%
  select(species, body_mass_g, body_mass_kg, bill_ratio, large_penguin) %>%
  head(8)
```

---

## Step 5: Pivot to long format

```{r pivot-longer}
# pivot_longer() converts the four measurement columns into two columns:
# "measurement" (the former column name) and "value" (the numeric reading).
# The long format is required when we want to plot or summarise all
# measurements in a single grouped operation without repeating code.
penguins_long <- penguins_mutated %>%
  pivot_longer(
    cols      = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to  = "measurement",
    values_to = "value"
  )

cat("Rows in long format:", nrow(penguins_long),
    "(= ", nrow(penguins_mutated), "rows ×", 4, "measurements)\n")

penguins_long %>%
  select(species, island, sex, measurement, value) %>%
  head(8)
```

---

## Step 6: Join with island metadata

```{r join}
# left_join keeps all penguin rows and adds environmental metadata
# from the island table. Using a left join (not inner) ensures we do not
# silently drop penguins if an island name is missing from the lookup.
penguins_joined <- penguins_mutated %>%
  left_join(metadata_islands, by = "island")

# Quick sanity check: row count must not change after the join
stopifnot(nrow(penguins_joined) == nrow(penguins_mutated))

penguins_joined %>%
  select(species, island, region, protected_area, average_temperature_c) %>%
  head(6)
```

---

## Step 7: Group, summarise, and order

```{r summarise}
# Grouping by species, island, sex, and protected area lets us see
# how body size differs across these biological and ecological strata.
# na.rm = TRUE is added defensively even though NAs were already removed —
# good practice for functions that do not propagate NAs by default.
penguins_summary <- penguins_joined %>%
  group_by(species, island, sex, protected_area) %>%
  summarise(
    n                = n(),
    mean_body_mass_g = mean(body_mass_g, na.rm = TRUE),
    median_flipper   = median(flipper_length_mm, na.rm = TRUE),
    mean_bill_ratio  = mean(bill_ratio, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_body_mass_g))

penguins_summary
```

**Key finding:** Gentoo males on Biscoe island are the heaviest group
(mean ≈ 5 480 g), followed by Gentoo females (≈ 4 640 g).
Both islands are protected areas, which may reflect lower disturbance levels.
Adelie penguins are consistently lighter across both islands regardless of sex.

---

## Summary of analytical outputs

| Object | Description |
|---|---|
| `penguins_cleaned_homework` | Standardised labels, no NAs in key columns |
| `penguins_filtered` | High/medium quality, Biscoe + Dream only, body mass > 3 000 g |
| `penguins_mutated` | Added `body_mass_kg`, `bill_ratio`, `large_penguin` |
| `penguins_long` | Long format via `pivot_longer()` — ready for faceted plots |
| `penguins_joined` | Island metadata appended via `left_join()` |
| `penguins_summary` | Final grouped summary table, ordered by mean body mass |
