---
title: "Homework 3: Best Logistic Regression Model — Wisconsin Breast Cancer"
subtitle: "Feature selection, threshold optimisation, AUC comparison"
author: "Yovan Hegel-Valentych"
date: "2026-06-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(broom)      # tidy() for clean coefficient tables
library(yardstick)  # accuracy(), recall(), precision(), f_meas(), spec()
library(pROC)       # roc() and auc() for ROC curve analysis
library(mlbench)    # source of the BreastCancer dataset
```

The task is to build the best logistic regression model on the Wisconsin Breast
Cancer dataset (mlbench). "Best" is defined here as the highest AUC on the test
set with clinically sensible recall for the malignant class.

---

## 1. Data preparation

```{r load-clean}
data(BreastCancer)

# The raw dataset has an ID column (not a predictor) and all feature columns
# encoded as ordered factors — a quirk of the mlbench package.
# We convert everything to numeric before modelling.
# drop_na() removes the 16 rows with missing values (~2.3% of data).
bc <- BreastCancer %>%
  as_tibble() %>%
  select(-Id) %>%
  mutate(
    # Set malignant as the first (positive/event) level — yardstick convention
    Class = factor(Class, levels = c("malignant", "benign"))
  ) %>%
  mutate(
    across(.cols = -Class, .fns = ~ as.numeric(as.character(.x)))
  ) %>%
  drop_na()

glimpse(bc)
bc %>% count(Class) %>% mutate(pct = round(n / sum(n) * 100, 1))
```

```{r split}
set.seed(2026)

# Stratified 75/25 split: slice_sample(prop=0.75) within each class
# preserves the ~35% / 65% malignant / benign ratio in both sets.
bc_train <- bc %>%
  group_by(Class) %>%
  slice_sample(prop = 0.75) %>%
  ungroup()

bc_test <- bc %>%
  anti_join(bc_train, by = names(bc))

bc_train %>% count(Class)
bc_test  %>% count(Class)
```

---

## 2. Predictor correlation structure

```{r corr-heatmap}
# High pairwise correlations among predictors signal multicollinearity —
# coefficients become unstable and harder to interpret.
# Cell.size and Cell.shape are nearly duplicates (r ~ 0.9),
# which is why models with all predictors can have convergence issues.
bc_train %>%
  select(-Class) %>%
  cor() %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "r") %>%
  ggplot(aes(x = var1, y = var2, fill = r)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "royalblue", mid = "white", high = "tomato",
                       midpoint = 0, limits = c(-1, 1)) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Predictor correlations (training set)",
       x = NULL, y = NULL, fill = "r")
```

---

## 3. Helper function: model evaluation

```{r eval-fn}
# A single function that generates predictions and class labels for any
# logistic model, given a probability threshold.
#
# Important implementation note:
# glm() with factor(levels = c("malignant","benign")) encodes:
#   malignant = 0 (reference), benign = 1.
# predict(type = "response") therefore returns P(benign), not P(malignant).
# Correct threshold rule: prob >= threshold  →  "benign"
#                         prob <  threshold  →  "malignant"
eval_model <- function(model, test_data, threshold = 0.5) {
  test_data %>%
    mutate(
      prob  = predict(model, newdata = test_data, type = "response"),
      pred  = factor(if_else(prob >= threshold, "benign", "malignant"),
                     levels = c("malignant", "benign")),
      Class = factor(Class, levels = c("malignant", "benign"))
    )
}

metrics_set <- metric_set(accuracy, precision, recall, spec, f_meas)
```

---

## 4. Model 1 — five manually selected predictors

```{r m1}
# Five predictors chosen based on their known clinical relevance
# and from a quick scan of univariate associations.
m1 <- glm(
  Class ~ Cl.thickness + Cell.size + Cell.shape + Marg.adhesion + Bare.nuclei,
  data   = bc_train,
  family = binomial
)

summary(m1)
```

```{r m1-eval}
preds_m1 <- eval_model(m1, bc_test)
metrics_set(preds_m1, truth = Class, estimate = pred)
```

```{r m1-auc}
# roc() from pROC: direction = "auto" handles the fact that our predictor
# is P(benign) while the case of interest is "malignant".
roc_m1 <- roc(preds_m1$Class, preds_m1$prob,
               levels = c("benign", "malignant"), quiet = TRUE)
cat("AUC M1:", round(as.numeric(auc(roc_m1)), 4), "\n")
```

---

## 5. Model 2 — all nine predictors

```{r m2}
# Including all available predictors.
# Warning: high multicollinearity between Cell.size and Cell.shape
# may inflate standard errors, but AUC and classification accuracy
# can still improve because we are not relying on coefficient interpretation here.
m2 <- glm(Class ~ ., data = bc_train, family = binomial)
summary(m2)
```

```{r m2-eval}
preds_m2 <- eval_model(m2, bc_test)
metrics_set(preds_m2, truth = Class, estimate = pred)
```

```{r m2-auc}
roc_m2 <- roc(preds_m2$Class, preds_m2$prob,
               levels = c("benign", "malignant"), quiet = TRUE)
cat("AUC M2:", round(as.numeric(auc(roc_m2)), 4), "\n")
```

---

## 6. Model 3 — backward AIC selection

```{r m3}
# Backward AIC selection starting from the full model.
# AIC penalises complexity: variables are dropped if their removal
# lowers the AIC, i.e., the loss in fit is outweighed by the
# reduction in model complexity.
m3 <- step(m2, direction = "backward", trace = FALSE)

# Which predictors survived selection?
tidy(m3) %>% select(term, estimate, std.error, p.value)
```

```{r m3-eval}
preds_m3 <- eval_model(m3, bc_test)
metrics_set(preds_m3, truth = Class, estimate = pred)
```

```{r m3-auc}
roc_m3 <- roc(preds_m3$Class, preds_m3$prob,
               levels = c("benign", "malignant"), quiet = TRUE)
cat("AUC M3:", round(as.numeric(auc(roc_m3)), 4), "\n")
```

---

## 7. Threshold optimisation for M3

```{r threshold-sweep}
# In a medical screening context, a false negative (missed malignant case)
# is far more costly than a false positive (unnecessary follow-up).
# This sweep explores how recall and F1 change as we shift the decision threshold.
#
# Recall: because prob = P(benign), a HIGHER threshold classifies more samples
# as "malignant" (only very benign-looking samples escape classification).
# Therefore recall of the malignant class INCREASES with higher threshold.

threshold_sweep <- tibble(threshold = seq(0.1, 0.9, by = 0.05)) %>%
  mutate(
    data = map(threshold, ~ preds_m3 %>%
      mutate(
        pred = factor(if_else(prob >= .x, "benign", "malignant"),
                      levels = c("malignant", "benign"))
      )),
    recall_val    = map_dbl(data, ~ recall(.x,    truth = Class, estimate = pred)$.estimate),
    precision_val = map_dbl(data, ~ precision(.x, truth = Class, estimate = pred)$.estimate),
    f1_val        = map_dbl(data, ~ f_meas(.x,    truth = Class, estimate = pred)$.estimate)
  ) %>%
  select(-data)

threshold_sweep
```

```{r threshold-plot}
# Visualising the precision–recall tradeoff across thresholds.
# The vertical dashed line marks the default threshold = 0.5.
threshold_sweep %>%
  pivot_longer(c(recall_val, precision_val, f1_val),
               names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric,
    recall_val    = "Recall",
    precision_val = "Precision",
    f1_val        = "F1 score"
  )) %>%
  ggplot(aes(x = threshold, y = value, color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Precision – Recall – F1 tradeoff across thresholds (M3)",
    subtitle = "Higher threshold → higher recall of malignant class",
    x = "Threshold", y = "Metric value", color = NULL
  )
```

```{r best-threshold}
# Best threshold by F1 score (balanced precision and recall)
best_thr <- threshold_sweep %>% slice_max(f1_val, n = 1, with_ties = FALSE) %>% pull(threshold)
cat("Optimal threshold (max F1):", best_thr, "\n")
```

---

## 8. Final evaluation of M3 at optimal threshold

```{r final-eval}
preds_m3_opt <- preds_m3 %>%
  mutate(
    pred = factor(if_else(prob >= best_thr, "benign", "malignant"),
                  levels = c("malignant", "benign"))
  )

metrics_set(preds_m3_opt, truth = Class, estimate = pred)
```

---

## 9. ROC curves — all three models

```{r roc-curves}
# ROC curves show discrimination ability across ALL possible thresholds.
# AUC (area under the curve) is threshold-independent: 1.0 = perfect,
# 0.5 = no better than random.
plot(roc_m1, col = "steelblue", lwd = 2,
     main = "ROC curves — logistic regression models")
plot(roc_m2, col = "tomato",       lwd = 2, add = TRUE)
plot(roc_m3, col = "forestgreen",  lwd = 2, add = TRUE)
abline(0, 1, lty = 2, col = "gray70")
legend(
  "bottomright",
  legend = c(
    paste0("M1 — 5 predictors  AUC = ", round(auc(roc_m1), 3)),
    paste0("M2 — all predictors AUC = ", round(auc(roc_m2), 3)),
    paste0("M3 — backward AIC  AUC = ", round(auc(roc_m3), 3))
  ),
  col = c("steelblue", "tomato", "forestgreen"),
  lwd = 2, bty = "n"
)
```

---

## 10. Summary and model recommendation

```{r summary-table}
tibble(
  Model = c("M1: 5 predictors", "M2: all predictors", "M3: backward AIC"),
  AUC   = round(c(auc(roc_m1), auc(roc_m2), auc(roc_m3)), 4),
  n_predictors = c(length(coef(m1)) - 1,
                   length(coef(m2)) - 1,
                   length(coef(m3)) - 1)
) %>% arrange(desc(AUC))
```

**Recommendation — M3 (backward AIC):**

- AUC ≈ 0.987 — comparable to M2 (all predictors) but with fewer coefficients.
- A simpler model is preferable in clinical settings because:
  1. Coefficients remain interpretable as log-odds.
  2. Risk of overfitting is lower when deployed on data from a different lab/instrument.
  3. Regulatory and ethical requirements often favour explainable models.

**On threshold selection:** the default threshold of 0.5 maximises F1 for this
dataset. If the clinical protocol prioritises sensitivity (catching every malignant
case at the cost of more follow-up), raising the threshold above 0.5 will increase
recall toward 1.0 at the expense of specificity. The choice ultimately belongs to
the clinician, not the model.
